---
title: "amptk_fungalpipeline"
author: "Claire Willing"
date: "2024-01-03"
output: html_document
bibliography: references.bib
---

## Bioinfomatic steps in AMPtk

**NOTE:** AMPtk is not working with M1 or M2 chips (latest Mac computers); you are able to run it, but Claire has had to run through Docker in order to use it eg (amptk-docker illumina -i SeqDataTest -o TerracesTest -f CTTGGTCATTTAGAGGAAGTAA -r GCTGCGTTCTTCATCGATGC --require_primer off --rescue_forward off --primer_mismatch 2)

**On a Mac:**

```{bash echo=FALSE}
brew install --docker
```

## Set up

### Step 1: process reads; sub in your primers

```{zsh engine.opts='-l'}
export ITS5_8S_FUN=AACTTTYRRCAAYGGATCWCT
export ITS2_FUN=AGCCTCCGCTTATTGATATGCTTAART
export ITS1F=CTTGGTCATTTAGAGGAAGTAA
export ITS2=GCTGCGTTCTTCATCGATGC

amptk-docker illumina \
  -i data/raw_data/sequencing \
  -o trimmed \
  -f $ITS5_8S_FUN -r $ITS2_FUN \
  --require_primer on \
  --rescue_forward off \
  --primer_mismatch 6  \
  --cleanup
```

## Clustering data; 2 errors in alignment allowed (longer seq length); default of 97% similarity

```{zsh, engine.opts='-l'}
amptk-docker cluster -i trimmed.demux.fq.gz -e 1.0 -o clustered   
```

## Filter table based on what is in negative controls

```{zsh, engine.opts='-l'}
amptk-docker filter -i clustered.otu_table.txt -f clustered.cluster.otus.fa -p 0.005 # Set p-value threshold
```

## Assign taxonomy (used ITS database)

Change the -o STUDY to your project name eg Westside

```{zsh, engine.opts='-l'}
amptk-docker taxonomy -f clustered.cluster.otus.fa -i clustered.otu_table.txt -m trimmed.mapping_file.txt -d ITS -o STUDY
```

### Reorganize files for clarity;

```{zsh, engine.opts='-l'}
# Create necessary directories
mkdir -p data/processed_data/
mkdir -p data/intermediate_data/amptk/logs
mkdir -p data/intermediate_data/amptk/otu_tables

mv -v trimm* cluster* STUDY* data/intermediate_data/amptk/ 2>/dev/null

mv -v data/intermediate_data/amptk/*.log data/intermediate_data/amptk/logs/ 2>/dev/null

mv -v data/intermediate_data/amptk/STUDY* data/intermediate_data/amptk/otu_tables/ 2>/dev/null

```

## NOW IN R

### Read in libraries

```{r message=FALSE, warning=FALSE}
# Function to install and load packages
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      if (pkg == "phyloseq") {
        if (!requireNamespace("BiocManager", quietly = TRUE)) {
          install.packages("BiocManager")
        }
        BiocManager::install(pkg)
      } else {
        install.packages(pkg)
      }
    }
    library(pkg, character.only = TRUE)
  }
}

# List of required packages
required_packages <- c(
  "phyloseq",  # For working with sequencing data
  "dplyr",     # For data manipulation
  "ggplot2"    # For plotting
)

# Install and load packages
install_and_load(required_packages)
```

# Read in Seq data

#### Create phyloseq object with otu table

```{r}
# Set file paths for BIOM and mapping files
biom_file <- "data/intermediate_data/amptk/otu_tables/STUDY.biom"

# Import BIOM and metadata
BIOM <- import_biom(biom_file, parseFunction = parse_taxonomy_greengenes)

#pull sample names to match with your metadata
biom_sample_names <- sample_names(BIOM)

map_file <- data.frame(SampleNames = biom_sample_names, row.names = biom_sample_names)

META <- sample_data(map_file)

Phyloseq <- merge_phyloseq(BIOM, META)
```

#### Load Funfun database

```{r}
devtools::install_github("ropenscilabs/datastorr")
devtools::install_github("traitecoevo/fungaltraits")
```

## Merge with Funfun and fungaltraits for downstream analysis [@flores-moreno; @põlme2020]

```{r}
library(fungaltraits)
Funfun <- fungal_traits()
fungaltraits <- read.csv("fungaltraitsDB/FungalTraits 1.2_ver_16Dec_2020 - V.1.2.csv") %>% rename(Genus = GENUS)

Phylogeny<-c("Phylum", "Class", "Order", "Family")

tax_data <- tax_table(Phyloseq) %>%
  as.data.frame() %>%
  mutate(OTU = rownames(.)) %>% 
  left_join(select(fungaltraits, -Phylogeny), 
                   by = "Genus") %>%
  mutate(speciesMatched = Species) %>%
  left_join(select(Funfun, -Genus), 
            by = "speciesMatched")%>%
  mutate(OTU = make.unique(as.character(OTU))) 

rownames(tax_data) <- tax_data$OTU

tax_table(Phyloseq) <- tax_data %>%
  select(-OTU) %>%
  as.matrix()
```

### Store and save your phyloseq object

```{r}
# Rename STUDY with your project name
saveRDS(Phyloseq, file = "data/processed_data/STUDY_otu_table.rds")
```
